A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-22 text/conclusion.tex
Spellcheck (except proof.tex)

We have proposed a new multi-stage calculus \LMD with dependent types.
\LMD enables finer-grained typing of multi-stage programming comparing to existing type systems.
This extention prevents some bugs relating to generated functions with multi-stage programming,
such as \verb|vadd| function for vectors of a fixed length.
We proved properties of \LMD including preservation, confluence, 
strong normalization for full reduction, and progress for staged reduction.
The technical difficulty of extention is how to deal with the symbol of CSP $\%$ in the equation rules.
We resolve this problem by ignoring $\%$ under some condition.

>>>>

We have proposed a new multi-stage calculus \LMD with dependent types.
\LMD enables finer-grained typing of multi-stage programming comparing to existing type systems.
This extension prevents some bugs relating to generated functions with multi-stage programming,
such as \verb|vadd| function for vectors of a fixed length.
We proved properties of \LMD including preservation, confluence, 
strong normalization for full reduction, and progress for staged reduction.
The technical difficulty of extension is how to deal with the symbol of CSP $\%$ in the equation rules.
We resolve this problem by ignoring $\%$ under some condition.

### 2019-06-22 text/formal-defnition.tex
Spellcheck (except proof.tex)

The rules for kind wellformedness and kiding is straightforward
adaptation from \LLF and \LTP, except for the rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

Since the syntax of kinds, types, and terms is mutually recursive,
corresponding notions of equivalence are also mutually recursive.
They are congruences closed under a few axioms for term equivalence.
So, the rules for kind and type equivalences are not very interesting
except implict CSP is allowed.
We show a few representative rules below.

@@

This rule is motivated by a practical consideration.  We will discuss
the motivation through an example of specializing a vector addition
function with respect to the size of vectors.  The function
vadd (with an auxilirary function vadd$_1$) below implements a code
generator in a hypothetical language based on \LTP.  This language
provides definitions by \textbf{let}, recursive functions (represented
by \textbf{fix}), \textbf{if}-expressions, and primitives cons, head,
and tail to maniputlate vectors.
We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

@@

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation context and of the hole, respectively.  The grammer represents that staged reduction is left-to-right and call-by-value and terms under \(\Lambda\) is reduced.
A few examples of evaluation context are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \epsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

@@

This reduction relation represents a term reduces in a deterministic,
left-to-right, call-by-value manner.  An application of an abstraction
is executed only at stage \(\varepsilon\) and only a quotation at
stage \(\varepsilon\) is spiliced into the surrounding code---notice
that, if \(\TB_\alpha v^\alpha\) is at stage \(\varepsilon\), then the
redex \(\TBL_\alpha \TB_\alpha v^\alpha\) is at stage \(\alpha\).
In other words, terms in brackets are not evaluated until the terms are run
and arguments of a function are evaluated before the application.
We show an example of staged reduction.
Underlines show the redexes.
\begin{align*}
                    & (\Lambda\alpha.(\TB_\alpha \underline{\TBL_\alpha \TB_\alpha ((\lambda x:\I.x)\ 10))})\ \varepsilon \\
  \longrightarrow_s & \underline{(\Lambda\alpha.(\TB_\alpha ((\lambda x:\I.x)\ 10)))\ \varepsilon}                        \\
  \longrightarrow_s & \underline{(\lambda x:\I.x)\ 10}                                                                    \\
  \longrightarrow_s & 10
\end{align*}

>>>>

The rules for kind wellformedness and kinding is straightforward
adaptation from \LLF and \LTP, except for the rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

Since the syntax of kinds, types, and terms is mutually recursive,
corresponding notions of equivalence are also mutually recursive.
They are congruences closed under a few axioms for term equivalence.
So, the rules for kind and type equivalences are not very interesting
except implicit CSP is allowed.
We show a few representative rules below.

@@

This rule is motivated by a practical consideration.  We will discuss
the motivation through an example of specializing a vector addition
function with respect to the size of vectors.  The function
vadd (with an auxiliary function vadd$_1$) below implements a code
generator in a hypothetical language based on \LTP.  This language
provides definitions by \textbf{let}, recursive functions (represented
by \textbf{fix}), \textbf{if}-expressions, and primitives cons, head,
and tail to manipulate vectors.
We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

@@

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation context and of the hole, respectively.  The grammar represents that staged reduction is left-to-right and call-by-value and terms under \(\Lambda\) is reduced.
A few examples of evaluation context are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \epsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

@@

This reduction relation represents a term reduces in a deterministic,
left-to-right, call-by-value manner.  An application of an abstraction
is executed only at stage \(\varepsilon\) and only a quotation at
stage \(\varepsilon\) is spliced into the surrounding code---notice
that, if \(\TB_\alpha v^\alpha\) is at stage \(\varepsilon\), then the
redex \(\TBL_\alpha \TB_\alpha v^\alpha\) is at stage \(\alpha\).
In other words, terms in brackets are not evaluated until the terms are run
and arguments of a function are evaluated before the application.
We show an example of staged reduction.
Underlines show the redexes.
\begin{align*}
                    & (\Lambda\alpha.(\TB_\alpha \underline{\TBL_\alpha \TB_\alpha ((\lambda x:\I.x)\ 10))})\ \varepsilon \\
  \longrightarrow_s & \underline{(\Lambda\alpha.(\TB_\alpha ((\lambda x:\I.x)\ 10)))\ \varepsilon}                        \\
  \longrightarrow_s & \underline{(\lambda x:\I.x)\ 10}                                                                    \\
  \longrightarrow_s & 10
\end{align*}

### 2019-06-22 text/introduction.tex
Revise.
Minor edits.
Spellcheck (except proof.tex)

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specilization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

\AI{Informal discussion on how important CSP is?}
\AI{Comments on Appendix required.}

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.  Section
\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.

>>>>

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specialization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.  Section
\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.  For reference, we give the full definition of \LMD
\AI{and proofs of properties}.

### 2019-06-22 text/properties.tex
Reduce the number of pages.
Spellcheck (except proof.tex)

\red{TODO: Fix the proof.}
\red{$\G\V J@A$における代入を定義する必要があるか?}

@@

Strong Normalization is also an important property, which guarantees that
no typed term has an infinite reduction sequence.
We prove this thorem by translating \LMD to the symply typed lambda calculus~\cite{harper1993framework}.
\AI{what is a standard citation for a proof of SN of a dependently typed calculus?}.
\red{とりあえずA Framework for Defining Logicsを引用しました。}

@@

\begin{proof}
	In order to prove this theorem, we define a translation $(\cdot)^\natural$ from \LMD\ to the symply typed lambda calculus.
	Second, we prove the $\natural$-translation preserves typing and reduction.
	Then, we can prove Strong Normalization of \LMD from Strong Normalization of the symply typed lambda calculus.
\end{proof}

@@

Progress is a collorary of Unique Decomposition.

>>>>

@@

Strong Normalization is also an important property, which guarantees that
no typed term has an infinite reduction sequence.
We prove this theorem by translating \LMD to the simply typed lambda calculus~\cite{harper1993framework}.

@@

\begin{proof}
	In order to prove this theorem, we define a translation $(\cdot)^\natural$ from \LMD\ to the simply typed lambda calculus.
	Second, we prove the $\natural$-translation preserves typing and reduction.
	Then, we can prove Strong Normalization of \LMD from Strong Normalization of the simply typed lambda calculus.
\end{proof}

@@

Progress is a corollary of Unique Decomposition.

### 2019-06-22 text/related-work.tex
Spellcheck (except proof.tex)

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionstic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

@@

Although there are studies on combinations of multi-stage programming and other programming features such as mutable cells~\cite{kiselyov2016refined},
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging},
a combination with dependent types has been little studied.
One exception is Brady and Hammond~\cite{brady2006dependently},
who have discussed a
combination of dependently typeed programming with staging in the
style of MetaOCaml to implement a staged interpreter, which is
statically guaranteed to generate well typed code.  However, they
focused on concrete programming examples and there is no theoretical investigation
of the programming language they used.

@@

% Although we define type equivalence of \LMD with composition of equivalence rules,
% there is another candiate to define,
% which gives reduction on types and compare the results of reduction such as~\cite{sorensen2006lectures}.
% This method is better than one of \LMD because equivalence rules become simple.
% However, we reject it because it cannot handle CSP flexibly.

>>>>

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionistic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

@@

Although there are studies on combinations of multi-stage programming and other programming features such as mutable cells~\cite{kiselyov2016refined},
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging},
a combination with dependent types has been little studied.
One exception is Brady and Hammond~\cite{brady2006dependently},
who have discussed a
combination of dependently typed programming with staging in the
style of MetaOCaml to implement a staged interpreter, which is
statically guaranteed to generate well typed code.  However, they
focused on concrete programming examples and there is no theoretical investigation
of the programming language they used.

@@

% Although we define type equivalence of \LMD with composition of equivalence rules,
% there is another candidate to define,
% which gives reduction on types and compare the results of reduction such as~\cite{sorensen2006lectures}.
% This method is better than one of \LMD because equivalence rules become simple.
% However, we reject it because it cannot handle CSP flexibly.
